---
title: "In-Class Activity: Multiple Regression and Inference"
output: 
  html_document:
    toc: yes
    toc_depth: 2
    df_print: kable
    highlight: tango
    includes:
      in_header: wcag-overrides.html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Learning goals

By the end of this activity, you should be able to:

- interpret a coefficient in a multiple regression using “holding other included variables constant”;
- explain how omitted-variable bias can change a coefficient;
- compare $R^2$ and adjusted $R^2$;
- conduct individual t-tests and construct confidence intervals;
- write a joint null hypothesis;
- distinguish restricted and unrestricted models; and
- calculate and interpret an F-test for joint significance.


# 1. Set up and explore the data

We will use `wage1` from the `wooldridge` package. Each observation represents one worker. The variables of interest are:

- `wage`: average hourly earnings;
- `lwage`: natural log of hourly earnings;
- `educ`: years of education;
- `exper`: years of potential labor-market experience;
- `tenure`: years with the current employer; and
- `female`: 1 if the worker is a woman and 0 otherwise.

```{r packages}
library(wooldridge)
library(tidyverse)
library(knitr)
```

```{r}
source('reporting.R')
```

```{r inspect-data}
data("wage1")

# Open the documentation and inspect the main variables of interest
?wage1

# Check sample size, missing values, and descriptive statistics
#summary(wage1)
mk_dataset_summary(wage1)
```

- `wage`: average hourly earnings;
- `lwage`: natural log of hourly earnings;
- `educ`: years of education;
- `exper`: years of potential labor-market experience;
- `tenure`: years with the current employer; and
- `female`: 1 if the worker is a woman and 0 otherwise.

1. What is the unit of observation? Which variables are continuous, and which is binary?
wage is dollars, lwage is the natural log of dollars. educ and exper and tenure are years. female is a binary variable. wage / lwage / educ/ exper/ tenure are all quantitative. female is binary.


2. Before estimating anything, predict the sign of each slope coefficient in


$$
lwage_i=\beta_0+\beta_1 educ_i+\beta_2 exper_i+
\beta_3 tenure_i+\beta_4 female_i+u_i.
$$
$\beta_1 > 0$
$\beta_2 > 0$
$\beta_3 > 0$
$\beta_4 < 0$



# 2. Visualize selected relationships

Create two plots:

1. a scatterplot of `educ` and `lwage` with a fitted line; and
2. a boxplot of `lwage` by gender.

Include informative labels and alt text.

```{r education-plot, fig.alt="Scatterplot of log hourly earnings against years of education with a fitted linear regression line."}
# Education and log-wage plot
ggplot(wage1, aes(x = educ, y = lwage)) + 
  geom_point(color = "steelblue") + 
  geom_smooth(method = 'lm', se = FALSE, formula = y ~ x, color = "darkorchid")
```


```{r gender-plot, fig.alt="Side-by-side boxplots comparing log hourly earnings for men and women in the wage sample."}
wage1 <-  wage1 %>%
  mutate(gender = factor(female, levels = c(0, 1), labels = c("men", "women")))
# Create a labeled gender factor and make the boxplot
ggplot(wage1, aes(x = gender, y = lwage, fill = gender)) +
  geom_boxplot(show.legend = FALSE) 

```

3. What do the plots suggest about the relationships? Why can neither plot show the partial relationship between a predictor and earnings?

The scatterplot shows that education is positively associated with higher log wages in the sample. The boxplot shows us that women have a lower lwage in the sample on average. These are unconditional relationships (only looking at one variable vs. lwage), these plots aren't including any other variable, aren't holding other factors constant.



# 3. From simple to multiple regression

Estimate these models:

$$
\text{Model 1: } lwage_i=\beta_0+\beta_1 educ_i+u_i
$$

$$
\text{Model 2: } lwage_i=\beta_0+\beta_1 educ_i+\beta_2 exper_i+
\beta_3 tenure_i+\beta_4 female_i+u_i.
$$

```{r estimate-models}
# Estimate the simple model as model_1
model_1 <- lm(lwage ~ educ, data = wage1)

# Estimate the multiple regression as model_2

model_2 <- lm(lwage ~ educ + exper + tenure + female, data = wage1)

# Display both summaries
#summary(model_1)
#summary(model_2)

```

```{r, results = "asis"}
mk_coeftab(model_1, caption = "model 1 summary")

```

```{r, results = "asis"}
mk_coeftab(model_2, caption = "model 2 summary")

```

4. Write the fitted equation for Model 2.

$\hat{lwage} = 0.5 + 0.087 educ + 0.005 exper + 0.017 tenure - 0.3 female$

5. Interpret the coefficients on `educ`, `tenure`, and `female`. Use exact percentage transformations where appropriate:
Each continuous slope is the earnings difference associated with a one-unit increase in that x, holding the other x variables constant. So for educ, and because the y variable is log transformed, and if we use the approximation, a one year increase in educ is associated with a 0.087 x 100 = 8.7% increase in wages.
Similar interpretation for tenure (its slope is the earnings difference associated with a one unit increase in tenure, holding the other variables constant). In this case, 0.017 x 100 = 1.7% increase in wages.
The female coefficient (because it's binary), it's the conditional log-wage difference between women and men with the same values for education/ experience/ tenure


Approximate change: multiply the coefficient by 100. A one unit increase in x is associated with a 100 x 100 % change in Y.

This doesn't work for larger changes and honestly is just an approximation anyway; the better choice is to do the: 

$$
100(e^{\widehat\beta}-1).
$$
Exponentiating to "undo" the log (our dependent variable is log transformed) and then subtract 1 and multiply by 100 to turn it into a percent change. 

if beta = .33:

approximation is a one unit change in x -> .33 x 100 = 33% change in y.

Exactly:

```{r}
(exp(.33)-1) * 100
```


```{r exact-percentages}
# Calculate exact percentages for educ, tenure, and female
(exp(coef(model_2)["educ"])-1) * 100
(exp(coef(model_2)["tenure"])-1) * 100
(exp(coef(model_2)["female"])-1) * 100
```

6. Compare the education coefficient across Models 1 and 2. Give a plausible omitted-variable-bias explanation for the change, but do not claim that the multiple-regression estimate is automatically causal.

```{r}
coef(model_1)["educ"]
coef(model_2)["educ"]
coef(model_1)["educ"] - coef(model_2)["educ"]
```
A coefficient change is consistent with OVB in the simple model but doesn't prove that model 2 has ever control variable and also doesn't identify a causal return to education and also doesn't actually prove OVB. It gives a clue that it exists.


7. For an omitted variable $Z$ to bias the education coefficient, what two conditions must hold?

1. Z has to affect Y (after controlling for X) so that its $\beta$ (from the true model) is not equal to zero.
2. Z is correlated with X, so that $Cov(X, Z) \neq 0$

$plim(\tilde{\beta}_1) = \beta_1 + \beta_2 \frac{Cov(X, Z)}{Var(X)}$

Where X is in the true model and also the estimated model, $\beta_1$ is its coefficient from the true model, $\tilde{\beta}_1$ is its coefficient from the estimated model without Z, $\beta_2$ is Z's coefficient (its coefficient from the true model):

$y_i = \beta_0 + \beta_1 X + \beta_2 Z + u_i$



# 4. Fitted values, residuals, and model fit

Create a prediction for a worker with 16 years of education, 10 years of experience, 5 years of tenure, and `female = 1`. Report predicted log wage and predicted hourly wage.

```{r prediction}
new_worker <- tibble(educ = 16, exper = 10, tenure = 5, female = 1)

# Add predicted lwage and wage
new_worker$predicted_lwage <- predict(model_2, newdata = new_worker)
new_worker$predicted_wage <- exp(new_worker$predicted_lwage)

new_worker <- tibble(
  educ = 16,
  exper = 10,
  tenure = 5,
  female = 1
) |>
  mutate(
    predicted_lwage = predict(model_2, newdata = pick(everything())),
    predicted_wage = exp(predicted_lwage)
  )

new_worker |>
  mutate(
    female = if_else(female == 1, "Female", "Male"),
    across(c(predicted_lwage, predicted_wage), \(x) round(x, 2))
  ) |>
  rename(
    "Education (years)" = educ,
    "Experience (years)" = exper,
    "Tenure (years)" = tenure,
    "Sex" = female,
    "Predicted log wage" = predicted_lwage,
    "Predicted wage ($/hour)" = predicted_wage
  ) |>
  knitr::kable(
    caption = "Predicted wage for a worker with 16 years of education, 10 years of experience, and 5 years of tenure.",
    align = c("r", "r", "r", "l", "r", "r")
  ) |>
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover"),
    full_width = FALSE
  )

```

8. Interpret the prediction. Why should you avoid describing it as the worker’s guaranteed wage?

This is the fitted conditional mean implied by the model, not an individual guarantee. The residual component and omitted determinants create variation around the prediction.



Compare $R^2$, adjusted $R^2$, residual standard error, AIC, and BIC.

```{r model-fit}
# Create a compact model-fit table
model_fit <- tibble(
  model = c("Education only", "Multiple regression"),
  r_squared = c(summary(model_1)$r.squared, summary(model_2)$r.squared),
  adjusted_r_squared = c(summary(model_1)$adj.r.squared,
                         summary(model_2)$adj.r.squared),
  residual_standard_error = c(sigma(model_1), sigma(model_2)),
  AIC = c(AIC(model_1), AIC(model_2)),
  BIC = c(BIC(model_1), BIC(model_2))
)


model_fit |>
  rename(
    "Model" = model,
    "R-squared" = r_squared,
    "Adjusted R-squared" = adjusted_r_squared,
    "Residual standard error" = residual_standard_error
  ) |>
  mutate(
    across(
      c(
        "R-squared",
        "Adjusted R-squared",
        "Residual standard error"
      ),
      \(x) round(x, 3)
    ),
    across(c(AIC, BIC), \(x) round(x, 1))
  ) |>
  knitr::kable(
    caption = "Comparison of model fit statistics for the education-only and multiple regression models.",
    align = c("l", "r", "r", "r", "r", "r")
  ) |>
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover"),
    full_width = FALSE
  )

```

9. Why must Model 2’s $R^2$ be **at least as large** as Model 1’s? Why might adjusted $R^2$ fall after adding a variable?

OLS can assign zero slopes to added variables, so adding regressors can't increase minimized RSS and can't lower $R^2$. Adjusted $R^2$ penalizes added regressors and can fall when their fit improvement is too small.

10. Does the model with the largest $R^2$ necessarily have the best causal specification? Explain.

No. Maximizing fit can encourage inappropriate controls, post-treatment variables, overfitting, and data mining. Causal specification depends on theory, timing, and the data-generating process.




# 5. Individual inference in multiple regression

For an individual coefficient,

$$
t=\frac{\widehat\beta_j-\beta_{j,0}}
{SE(\widehat\beta_j)}.
$$

Test $H_0:\beta_{tenure}=0$ against $H_1:\beta_{tenure}\neq0$ by extracting the estimate and conventional standard error from Model 2 and calculating the t-statistic and two-sided p-value.

```{r tenure-test}
tenure_estimate <- coef(model_2)["tenure"]
tenure_se <- coef(summary(model_2))["tenure", "Std. Error"]
tenure_df <- df.residual(model_2)
tenure_t <- tenure_estimate / tenure_se
tenure_p <- 2 * pt(-abs(tenure_t), df = tenure_df)


tenure_results <- tibble(
  term = "Tenure",
  estimate = unname(tenure_estimate),
  standard_error = unname(tenure_se),
  t_statistic = unname(tenure_t),
  degrees_of_freedom = tenure_df,
  p_value = unname(tenure_p)
)

tenure_results |>
  rename(
    "Term" = term,
    "Estimate" = estimate,
    "Standard error" = standard_error,
    "t statistic" = t_statistic,
    "Degrees of freedom" = degrees_of_freedom,
    "p-value" = p_value
  ) |>
  mutate(
    across(
      c("Estimate", "Standard error", "t statistic"),
      \(x) round(x, 3)
    ),
    "p-value" = format.pval(`p-value`, digits = 3, eps = 0.001)
  ) |>
  knitr::kable(
    caption = "Hypothesis test results for the tenure coefficient.",
    align = c("l", "r", "r", "r", "r", "r")
  ) |>
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover"),
    full_width = FALSE
  )
```

11. State the decision at $\alpha=0.05$ and interpret the result in context.

At 5%, `r ifelse(tenure_p <= .05, "reject", "fail to reject")` the null of no partial association between tenure and log wages. The coefficient is interpreted holding education, experience, and gender constant.


Construct 95% confidence intervals for every Model 2 coefficient.

```{r coefficient-cis}
# Use confint()
model_2_ci <- confint(model_2, level = 0.95)


model_2_ci |>
  as.data.frame() |>
  tibble::rownames_to_column("Coefficient") |>
  setNames(c(
    "Coefficient",
    "Lower bound",
    "Upper bound"
  )) |>
  mutate(
    across(
      c("Lower bound", "Upper bound"),
      \(x) round(x, 3)
    )
  ) |>
  knitr::kable(
    caption = "95% confidence intervals for the multiple regression coefficients.",
    align = c("l", "r", "r")
  ) |>
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover"),
    full_width = FALSE
  )
```

12. Interpret the interval for `educ`. Which individual null hypotheses $H_0:\beta_j=0$ are rejected at the 5% level?
Under repeated sampling, 95% of intervals produced this way would contain the population coefficient. A coefficient’s two-sided zero null is rejected at 5% exactly when its corresponding interval excludes zero.


# 6. Why an F-test?

Suppose we want to know whether labor-market history contributes to the model after controlling for education and gender. The joint hypotheses are

$$
H_0:\beta_{exper}=0 \text{ AND } \beta_{tenure}=0
$$

$$
H_1:\beta_{exper}\neq0 \text{ OR } \beta_{tenure}\neq0.
$$

13. How many restrictions are in the null? What does “OR” in the alternative mean?
There are q=2 restrictions. “OR” means the alternative is true if at least one of the two coefficients differs from zero, rejection doesn't require both to be nonzero.



14. Why not just conduct two separate t-tests at $\alpha=0.05$ and reject the joint null if either is significant?
Separate 5% tests inflate the probability of at least one false rejection and ignore covariance between the coefficient estimates. The F-test controls the size of the single joint test.


The **unrestricted model** includes all variables:

$$
lwage_i=\beta_0+\beta_1 educ_i+\beta_2 exper_i+
\beta_3 tenure_i+\beta_4 female_i+u_i.
$$

The **restricted model** imposes the null by removing `exper` and `tenure`:

$$
lwage_i=\beta_0+\beta_1 educ_i+\beta_4 female_i+u_i.
$$

15. Explain why Model 2 is unrestricted and the smaller model is restricted. Which model can never have the larger residual sum of squares?

Model 2 is unrestricted because it estimates `exper` and `tenure` freely. The restricted model imposes the joint null by setting both slopes to zero. The unrestricted model cannot have larger RSS because it contains the restricted model as a special case.



# 7. Calculate the partial F-test by hand

Use the residual sum of squares (SSR) form:

$$
F=\frac{(SSR_R-SSR_U)/q}{SSR_U/(n-k_U-1)}
\sim F(q,n-k_U-1),
$$

where:

- $SSR_R$ is the restricted-model residual sum of squares;
- $SSR_U$ is the unrestricted-model residual sum of squares;
- $q$ is the number of restrictions;
- $k_U$ is the number of slopes in the unrestricted model; and
- $n-k_U-1$ is the unrestricted model’s residual degrees of freedom.

```{r partial-f-test}
restricted_model <- lm(lwage ~ educ + female, data = wage1)
unrestricted_model <- model_2

RSS_R <- sum(resid(restricted_model)^2)
RSS_U <- sum(resid(unrestricted_model)^2)
q <- 2
df_U <- df.residual(unrestricted_model)

partial_F <- ((RSS_R - RSS_U) / q) / (RSS_U / df_U)
partial_p <- pf(partial_F, df1 = q, df2 = df_U, lower.tail = FALSE)
partial_critical <- qf(0.95, df1 = q, df2 = df_U)

partial_F_results <- tibble(
  test = "Joint significance of experience and tenure",
  f_statistic = partial_F,
  numerator_df = q,
  denominator_df = df_U,
  critical_value = partial_critical,
  p_value = partial_p
)

partial_F_results |>
  rename(
    "Test" = test,
    "F statistic" = f_statistic,
    "Numerator df" = numerator_df,
    "Denominator df" = denominator_df,
    "Critical value" = critical_value,
    "p-value" = p_value
  ) |>
  mutate(
    across(
      c("F statistic", "Critical value"),
      \(x) round(x, 3)
    ),
    "p-value" = format.pval(`p-value`, digits = 3, eps = 0.001)
  ) |>
  knitr::kable(
    caption = "Partial F-test of the joint significance of experience and tenure.",
    align = c("l", "r", "r", "r", "r", "r")
  ) |>
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover"),
    full_width = FALSE
  )
```

16. State the decision and conclusion in context. Be clear about what's held constant.
At 5%, `r ifelse(partial_p <= .05, "reject", "fail to reject")` the joint null that experience and tenure have no partial association with log wages after controlling for education and gender.


Check the calculation by comparing the nested models.

```{r verify-partial-f, results = 'asis"}
# Use anova(restricted_model, unrestricted_model)
mk_anova(restricted_model)
mk_anova(unrestricted_model)
```

17. Match every element of the ANOVA comparison to the hand calculation: $q$, $SSR_R-SSR_U$, $F$, and the p-value.

The ANOVA `Df` change is $q$; `Sum of Sq` is $RSS_R-RSS_U$. its F and p-value match the hand calculation.


18. If the partial F-test rejects, does that prove both coefficients are individually nonzero? Explain why/ why not.

No. Rejection means at least one restriction is false or the pair is jointly inconsistent with zero. It does not establish that both individual coefficients differ from zero.



# 8. The overall F-test

The F-statistic in `summary(model_2)` tests

$$
H_0:\beta_{educ}=\beta_{exper}=\beta_{tenure}=\beta_{female}=0.
$$

19. What is the restricted model for this test? Extract the overall F-statistic and its two degrees of freedom, then calculate its p-value.

The restricted model is intercept-only. The output tests all four slopes jointly.

```{r overall-f-test}
overall_f <- summary(model_2)$fstatistic

# Display F, numerator df, denominator df, and p-value

overall_F_results <- tibble(
  test = "Overall model significance",
  f_statistic = unname(overall_f["value"]),
  numerator_df = unname(overall_f["numdf"]),
  denominator_df = unname(overall_f["dendf"]),
  p_value = pf(
    unname(overall_f["value"]),
    df1 = unname(overall_f["numdf"]),
    df2 = unname(overall_f["dendf"]),
    lower.tail = FALSE
  )
)

overall_F_results |>
  rename(
    "Test" = test,
    "F statistic" = f_statistic,
    "Numerator df" = numerator_df,
    "Denominator df" = denominator_df,
    "p-value" = p_value
  ) |>
  mutate(
    "F statistic" = round(`F statistic`, 3),
    "p-value" = format.pval(`p-value`, digits = 3, eps = 0.001)
  ) |>
  knitr::kable(
    caption = "Overall F-test of model significance.",
    align = c("l", "r", "r", "r", "r")
  ) |>
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover"),
    full_width = FALSE
  )

```



20. How does the overall F-test answer a different question from the partial F-test in Section 7?

The overall test asks whether any included slope contributes relative to an intercept-only model. The partial test asks whether experience and tenure add explanatory power after education and gender are already included.



# 9. Multicollinearity and specification

Inspect correlations among the continuous regressors.

```{r correlations}
# Correlation matrix for educ, exper, and tenure
cor(wage1[, c("educ", "exper", "tenure")]) |>
  as.data.frame() |>
  tibble::rownames_to_column("Variable") |>
  dplyr::rename(
    "Education" = educ,
    "Experience" = exper,
    "Tenure" = tenure
  ) |>
  dplyr::mutate(
    Variable = dplyr::recode(
      Variable,
      educ = "Education",
      exper = "Experience",
      tenure = "Tenure"
    ),
    dplyr::across(dplyr::where(is.numeric), \(x) round(x, 3))
  ) |>
  knitr::kable(
    caption = "Correlation matrix for education, experience, and tenure.",
    align = c("l", "r", "r", "r")
  ) |>
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover"),
    full_width = FALSE
  )
```

21. What happens to standard errors when regressors don't have a lot of independent variation? Does multicollinearity bias OLS coefficients?

Strong collinearity increases coefficient variances and standard errors because little independent variation remains to isolate partial effects. It does not itself bias OLS coefficients.


22. A researcher keeps adding variables just because they increase $R^2$. Identify two problems with this approach.

Problems include overfitting, mechanically favoring irrelevant variables, increasing estimator variance, adding inappropriate controls, and confusing predictive fit with causal identification. $R^2$ never penalizes complexity.

# 10. Final check

In 5–7 sentences, summarize the Model 2 results. Include:

- one partial-effect interpretation;
- one individual inference result;
- the partial F-test conclusion;
- the overall F-test conclusion; and
- one limitation affecting causal interpretation.



