---
title: "In-Class Activity: Inference in Simple Linear Regression"
output: 
  html_document:
    toc: yes
    toc_depth: 2
    df_print: kable
    highlight: tango
    includes:
      in_header: wcag-overrides.html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Learning goals

By the end of this activity, you should be able to:

- explain why regression estimates vary across samples;
- interpret a regression with a binary explanatory variable;
- conduct hypothesis tests about zero and nonzero coefficient values;
- construct and interpret confidence intervals;
- connect a binary-regressor model to a difference in group means; and
- compare conventional and unequal-variance standard errors.


# 1. Set up and inspect the data

We will use `wage2` from the `wooldridge` package. Each observation represents one worker. Our focal variables are:

- `wage`: monthly earnings in dollars;
- `lwage`: the natural log of monthly earnings;
- `educ`: years of education; and
- `urban`: an indicator equal to 1 if the worker lives in an urban area and 0 otherwise.

```{r packages}
library(wooldridge)
library(tidyverse)
library(knitr)
```

```{r inspect-data}
data("wage2")

# Open the documentation and inspect the focal variables

# Check the sample size and missing values

```

1. What is the unit of observation? What are the units of `wage`, `lwage`, and `educ`?



2. Use a frequency table to verify how `urban` is coded. Which category is the reference group when `urban = 0`?

```{r inspect-urban}
# Create a frequency table for urban

```




# 2. Compare the two groups

Create a labeled factor called `urban_status`, then calculate the sample size, mean log earnings, standard deviation, and variance for each group.

```{r group-summary}
wage2 <- wage2 |>
  mutate(
    urban_status = # label 0 as "Non-urban" and 1 as "Urban"
  )

# Create the group summary

```

3. Which group has the larger mean log earnings? Calculate the raw difference as urban mean minus non-urban mean.




Create a boxplot comparing log earnings across the groups. Use informative labels, accessible colors, and alt text.

```{r boxplot, fig.alt="Side-by-side boxplots comparing log monthly earnings for workers living in non-urban and urban areas."}
# Create the boxplot here

```

4. Describe the centers, spreads, and overlap. Can this plot by itself establish a population difference?






# 3. Connect a binary regressor to group means

Consider the population model

$$
lwage_i=\beta_0+\beta_1 urban_i+u_i.
$$

Under the zero conditional mean assumption,

$$
E(lwage_i\mid urban_i=0)=\beta_0
$$

and

$$
E(lwage_i\mid urban_i=1)=\beta_0+\beta_1.
$$

So

$$
\beta_1=E(lwage_i\mid urban_i=1)-E(lwage_i\mid urban_i=0).
$$

5. Interpret $\beta_0$ and $\beta_1$ in this application.



```{r fit-model}
# Estimate lwage on urban and save the model as urban_model

# Display the model summary

```

6. Check that the intercept equals one group mean and the slope equals the difference in group means.

```{r verify-means}
# Calculate the fitted mean for urban = 0 and urban = 1 from the coefficients

# Compare them with the observed group means

```



7. Convert the slope from log points to an exact percentage difference using

$$
100(e^{\widehat\beta_1}-1).
$$

```{r percent-difference}
# Calculate the approximate and exact percentage differences

```

> **Interpretation:**



# 4. Construct a hypothesis test by hand

Test

$$
H_0:\beta_1=0
\qquad \text{against} \qquad
H_1:\beta_1\neq0.
$$

Assuming **equal population variances**, the pooled variance and conventional standard error are

$$
s_p^2=\frac{(n_1-1)s_1^2+(n_0-1)s_0^2}{n_1+n_0-2}
$$

and

$$
SE(\bar Y_1-\bar Y_0)=
\sqrt{s_p^2\left(\frac{1}{n_1}+\frac{1}{n_0}\right)}.
$$

```{r test-by-hand}
urban_group <- wage2 |> filter(urban == 1)
nonurban_group <- wage2 |> filter(urban == 0)

n_1 <- # urban sample size
n_0 <- # non-urban sample size
mean_1 <- # urban mean lwage
mean_0 <- # non-urban mean lwage
var_1 <- # urban variance of lwage
var_0 <- # non-urban variance of lwage

mean_difference <- # urban mean minus non-urban mean
pooled_variance <- # pooled variance formula
conventional_se <- # conventional standard error
degrees_freedom <- # pooled-test degrees of freedom
t_statistic <- # test H0: beta_1 = 0
p_value <- # two-sided p-value using pt()

c(estimate = mean_difference, standard_error = conventional_se,
  t = t_statistic, df = degrees_freedom, p_value = p_value)
```

8. At $\alpha=0.05$, what is your decision? State the conclusion in context without saying that you “accept” the null.



9. Explain the p-value in context. Begin with “If the null hypothesis were true...”




10. Compare your estimate, standard error, t-statistic, and p-value with `summary(urban_model)`.




# 5. Test nonzero null hypotheses

Test $H_0:\beta_1=0.10$ and $H_0:\beta_1=0.20$, using a two-sided alternative in each case.

```{r nonzero-nulls}
null_values <- c(0.10, 0.20)

# Calculate a t-statistic and two-sided p-value for each null value

```

11. What is your decision for each test at $\alpha=0.05$? Why can the same sample support different decisions for different null values?




# 6. Construct confidence intervals

```{r confidence-intervals}
# Find the 90% and 95% critical values with qt()

# Calculate both confidence intervals

# Display them in a small table

```

12. Which interval is wider, and why?




13. Interpret the 95% confidence interval. Then translate its endpoints into percentage differences.




14. Use the 95% interval to evaluate $H_0:\beta_1=0$, $H_0:\beta_1=0.10$, and $H_0:\beta_1=0.20$. 




```{r verify-ci}
# Check the 90% and 95% intervals with confint()

```

# 7. Conventional versus unequal-variance inference

An unequal-variance standard error for the group difference is

$$
SE_{unequal}=\sqrt{\frac{s_1^2}{n_1}+\frac{s_0^2}{n_0}}.
$$

```{r unequal-variance-test}
unequal_se <- # unequal-variance standard error
welch_df <- # Welch-Satterthwaite degrees of freedom
welch_t <- # test H0: beta_1 = 0
welch_p <- # two-sided p-value

c(conventional_se = conventional_se,
  unequal_variance_se = unequal_se,
  welch_t = welch_t, welch_df = welch_df,
  welch_p_value = welch_p)
```

15. Does the standard-error choice change the conclusion? Why should the method still be reported?



```{r verify-t-tests}
# Verify the pooled and Welch tests with t.test()

```



# 8. Statistical significance and causal language

16. A classmate writes: “Living in an urban area causes monthly earnings to increase.” Rewrite the claim so it is supported by this analysis.




17. Name two factors that could affect earnings and be systematically different between urban and non-urban workers. What would leaving them out mean for $E(u\mid urban)=0$?




18. Is statistical significance the same as practical importance? What additional context would help you evaluate the size/ magnitude?



# 9. Final synthesis

In 4–6 sentences, summarize the estimated difference, uncertainty, hypothesis-test result, and one important limitation. Use percentage language and avoid causal language.



# Optional extension: sampling variability

Draw 500 bootstrap samples, estimate the urban coefficient in each, and plot its sampling distribution. Compare its center and standard deviation with the original estimate and conventional standard error.

```{r bootstrap-extension, eval=FALSE}
set.seed(8300)

# Write the bootstrap simulation here

```

