---
title: "In-Class Activity: Multiple Regression and Inference"
output: 
  html_document:
    toc: yes
    toc_depth: 2
    df_print: kable
    highlight: tango
    includes:
      in_header: wcag-overrides.html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Learning goals

By the end of this activity, you should be able to:

- interpret a coefficient in a multiple regression using “holding other included variables constant”;
- explain how omitted-variable bias can change a coefficient;
- compare $R^2$ and adjusted $R^2$;
- conduct individual t-tests and construct confidence intervals;
- write a joint null hypothesis;
- distinguish restricted and unrestricted models; and
- calculate and interpret an F-test for joint significance.


# 1. Set up and explore the data

We will use `wage1` from the `wooldridge` package. Each observation represents one worker. The variables of interest are:

- `wage`: average hourly earnings;
- `lwage`: natural log of hourly earnings;
- `educ`: years of education;
- `exper`: years of potential labor-market experience;
- `tenure`: years with the current employer; and
- `female`: 1 if the worker is a woman and 0 otherwise.

```{r packages}
library(wooldridge)
library(tidyverse)
library(knitr)
```

```{r inspect-data}
data("wage1")

# Open the documentation and inspect the main variables of interest

# Check sample size, missing values, and descriptive statistics

```

1. What is the unit of observation? Which variables are continuous, and which is binary?



2. Before estimating anything, predict the sign of each slope coefficient in


$$
lwage_i=\beta_0+\beta_1 educ_i+\beta_2 exper_i+
\beta_3 tenure_i+\beta_4 female_i+u_i.
$$



# 2. Visualize selected relationships

Create two plots:

1. a scatterplot of `educ` and `lwage` with a fitted line; and
2. a boxplot of `lwage` by gender.

Include informative labels and alt text.

```{r education-plot, fig.alt="Scatterplot of log hourly earnings against years of education with a fitted linear regression line."}
# Education and log-wage plot

```


```{r gender-plot, fig.alt="Side-by-side boxplots comparing log hourly earnings for men and women in the wage sample."}
# Create a labeled gender factor and make the boxplot

```

3. What do the plots suggest about the relationships? Why can neither plot show the partial relationship between a predictor and earnings?





# 3. From simple to multiple regression

Estimate these models:

$$
\text{Model 1: } lwage_i=\beta_0+\beta_1 educ_i+u_i
$$

$$
\text{Model 2: } lwage_i=\beta_0+\beta_1 educ_i+\beta_2 exper_i+
\beta_3 tenure_i+\beta_4 female_i+u_i.
$$

```{r estimate-models}
# Estimate the simple model as model_1

# Estimate the multiple regression as model_2

# Display both summaries

```

4. Write the fitted equation for Model 2.



5. Interpret the coefficients on `educ`, `tenure`, and `female`. Use exact percentage transformations where appropriate:

$$
100(e^{\widehat\beta}-1).
$$

```{r exact-percentages}
# Calculate exact percentages for educ, tenure, and female

```





6. Compare the education coefficient across Models 1 and 2. Give a plausible omitted-variable-bias explanation for the change, but do not claim that the multiple-regression estimate is automatically causal.




7. For an omitted variable $Z$ to bias the education coefficient, what two conditions must hold?




# 4. Fitted values, residuals, and model fit

Create a prediction for a worker with 16 years of education, 10 years of experience, 5 years of tenure, and `female = 1`. Report predicted log wage and predicted hourly wage.

```{r prediction}
new_worker <- tibble(educ = 16, exper = 10, tenure = 5, female = 1)

# Add predicted lwage and wage

```

8. Interpret the prediction. Why should you avoid describing it as the worker’s guaranteed wage?




Compare $R^2$, adjusted $R^2$, residual standard error, AIC, and BIC.

```{r model-fit}
# Create a compact model-fit table

```

9. Why must Model 2’s $R^2$ be **at least as large** as Model 1’s? Why might adjusted $R^2$ fall after adding a variable?




10. Does the model with the largest $R^2$ necessarily have the best causal specification? Explain.






# 5. Individual inference in multiple regression

For an individual coefficient,

$$
t=\frac{\widehat\beta_j-\beta_{j,0}}
{SE(\widehat\beta_j)}.
$$

Test $H_0:\beta_{tenure}=0$ against $H_1:\beta_{tenure}\neq0$ by extracting the estimate and conventional standard error from Model 2 and calculating the t-statistic and two-sided p-value.

```{r tenure-test}
tenure_estimate <- # coefficient estimate
tenure_se <- # conventional standard error
tenure_df <- # residual degrees of freedom
tenure_t <- # t-statistic
tenure_p <- # two-sided p-value

c(estimate = tenure_estimate, standard_error = tenure_se,
  t = tenure_t, df = tenure_df, p_value = tenure_p)
```

11. State the decision at $\alpha=0.05$ and interpret the result in context.




Construct 95% confidence intervals for every Model 2 coefficient.

```{r coefficient-cis}
# Use confint()

```

12. Interpret the interval for `educ`. Which individual null hypotheses $H_0:\beta_j=0$ are rejected at the 5% level?



# 6. Why an F-test?

Suppose we want to know whether labor-market history contributes to the model after controlling for education and gender. The joint hypotheses are

$$
H_0:\beta_{exper}=0 \text{ AND } \beta_{tenure}=0
$$

$$
H_1:\beta_{exper}\neq0 \text{ OR } \beta_{tenure}\neq0.
$$

13. How many restrictions are in the null? What does “OR” in the alternative mean?




14. Why not just conduct two separate t-tests at $\alpha=0.05$ and reject the joint null if either is significant?



The **unrestricted model** includes all variables:

$$
lwage_i=\beta_0+\beta_1 educ_i+\beta_2 exper_i+
\beta_3 tenure_i+\beta_4 female_i+u_i.
$$

The **restricted model** imposes the null by removing `exper` and `tenure`:

$$
lwage_i=\beta_0+\beta_1 educ_i+\beta_4 female_i+u_i.
$$

15. Explain why Model 2 is unrestricted and the smaller model is restricted. Which model can never have the larger residual sum of squares?




# 7. Calculate the partial F-test by hand

Use the residual sum of squares (SSR) form:

$$
F=\frac{(SSR_R-SSR_U)/q}{SSR_U/(n-k_U-1)}
\sim F(q,n-k_U-1),
$$

where:

- $SSR_R$ is the restricted-model residual sum of squares;
- $SSR_U$ is the unrestricted-model residual sum of squares;
- $q$ is the number of restrictions;
- $k_U$ is the number of slopes in the unrestricted model; and
- $n-k_U-1$ is the unrestricted model’s residual degrees of freedom.

```{r partial-f-test}
restricted_model <- # estimate the model implied by H0
unrestricted_model <- model_2

SSR_R <- # restricted residual sum of squares
SSR_U <- # unrestricted residual sum of squares
q <- # number of restrictions
df_U <- # unrestricted residual degrees of freedom

partial_F <- # calculate the F-statistic
partial_p <- # right-tail p-value using pf()
partial_critical <- # 5% critical value using qf()

c(F = partial_F, numerator_df = q, denominator_df = df_U,
  critical_value = partial_critical, p_value = partial_p)
```

16. State the decision and conclusion in context. Be clear about what's held constant.



Check the calculation by comparing the nested models.

```{r verify-partial-f}
# Use anova(restricted_model, unrestricted_model)

```

17. Match every element of the ANOVA comparison to the hand calculation: $q$, $SSR_R-SSR_U$, $F$, and the p-value.




18. If the partial F-test rejects, does that prove both coefficients are individually nonzero? Explain why/ why not.





# 8. The overall F-test

The F-statistic in `summary(model_2)` tests

$$
H_0:\beta_{educ}=\beta_{exper}=\beta_{tenure}=\beta_{female}=0.
$$

19. What is the restricted model for this test? Extract the overall F-statistic and its two degrees of freedom, then calculate its p-value.

```{r overall-f-test}
overall_f <- summary(model_2)$fstatistic

# Display F, numerator df, denominator df, and p-value

```



20. How does the overall F-test answer a different question from the partial F-test in Section 7?




# 9. Multicollinearity and specification

Inspect correlations among the continuous regressors.

```{r correlations}
# Correlation matrix for educ, exper, and tenure

```

21. What happens to standard errors when regressors don't have a lot of independent variation? Does multicollinearity bias OLS coefficients?




22. A researcher keeps adding variables just because they increase $R^2$. Identify two problems with this approach.



# 10. Final check

In 5–7 sentences, summarize the Model 2 results. Include:

- one partial-effect interpretation;
- one individual inference result;
- the partial F-test conclusion;
- the overall F-test conclusion; and
- one limitation affecting causal interpretation.



