---
title: "ECON-3300"
output:
  html_document:
    toc: true
    toc_float: true
    highlight: tango
    includes:
      in_header: wcag-overrides.html
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Homework03

Grading Rubric (per question):  
2 points if complete and correct  
0 points otherwise 


## Retrieving your work

This and all future homework .Rmd files are written in Markdown. The .Rmd file can be rendered as an .html file for easier viewing. 

Once you have opened the .Rmd file, you may begin writing your answers beneath each problem. Do not remove any of the original problems from the file, because that makes it more difficult for the me to grade your assignment. 


## Submitting your work

When you upload your files to Canvas, you are to submit ***two*** files:

a. Your reproducible document file (.Rmd) which should be saved as homework##-ID.Rmd. For example, my homework01 file would be saved as homework01-jperrigan.Rmd.

b. Your rendered reproducible document file (.html) which should be saved as homework##-ID.html. For example, my homework01 file would be saved as homework01-jperrigan.html.

Your ID is your Canvas username. It is also the text before the "@" symbol in your UNO email address.

You have an unlimited number of submissions, but only the  latest submission will be viewed and graded. **Proper means that the file is correctly named and formatted. Latest means that the file is the most recent submission prior to the deadline. Remember the .Rmd file needs to render properly to .html before submitting.** 


***

All students are expected to complete all problems. If any problem asks you a particular question, be sure to answer it completely (with code, written sentences, or both).

***Do not change anything in this file above the double line. Doing so will prevent your homework from being graded.***

***
***

# Data and packages

We will use the `meap01` dataset from the `wooldridge` package. It contains data for 1,823 Michigan elementary schools. The variables used here are:

- `math4`: percentage of fourth-grade students rated satisfactory in mathematics
- `exppp`: expenditures per pupil, in dollars
- `lexppp`: natural logarithm of expenditures per pupil
- `lunch`: percentage of students eligible for free or reduced-price lunch
- `enroll`: school enrollment

```{r packages}
# Install the package once if needed:
# install.packages("wooldridge")

library(wooldridge)
data("meap01")
```

---

**#1**  

**Exploring the variables**

Using `meap01`:

1. Report the number of observations, the mean, standard deviation, minimum, and maximum of `math4` and `exppp`.
2. In 2–3 sentences, describe the units and the amount of variation in the two variables.

Save the table or object containing your descriptive statistics as `prob01_stats` and your written response as a string named `prob01`.

**Answer:**

```{r, problem01}
# prob01_stats <-
# prob01 <-
```

---

**#2** 

**Visualizing and quantifying the relationship**

1. Construct a scatterplot with `exppp` on the horizontal axis and `math4` on the vertical axis. Add a fitted linear regression line and informative axis labels.
2. In 2–3 sentences, describe the direction, strength, and shape of the relationship. Note any observations that might have unusually high influence. Is there a high correlation by sight? What about when you calculate the correlation?

Save your written response as `prob02`.

**Answer:**

```{r, problem02}
# put your plot code here
# prob02 <-
```

---

**#3**  


Consider the population regression model

$$
math4_i = \beta_0 + \beta_1 exppp_i + u_i.
$$

In 4–6 sentences:

1. Explain the economic meaning of $\beta_0$ and $\beta_1$, including their units. Are they population parameters or sample estimates?
2. Explain what the error term $u_i$ contains. Give two concrete examples of factors that could belong in $u_i$.

Save your response as a string named `prob03`.

**Answer:**

```{r, problem03}
# prob03 <-
```

---

**#4** 

**Calculating the OLS estimators by hand**

Let $x_i=exppp_i$ and $y_i=math4_i$.

1. Compute the sample means $\bar{x}$ and $\bar{y}$.
2. Create the centered variables $x_i^c=x_i-\bar{x}$ and $y_i^c=y_i-\bar{y}$; save them as `xc` and `yc`.
3. Verify numerically that $\sum_i x_i^c=0$ and $\sum_i y_i^c=0$.
4. Calculate

$$
\hat\beta_1=\frac{\sum_i(x_i-\bar{x})(y_i-\bar{y})}{\sum_i(x_i-\bar{x})^2}
$$

and

$$
\hat\beta_0=\bar{y}-\hat\beta_1\bar{x}.
$$

Save the estimates as `b1_hat` and `b0_hat`. Do not use `lm()` in this problem.

**Answer:**

```{r, problem04}
# xc <-
# yc <-
# b1_hat <-
# b0_hat <-
```

---

**#5**  

Estimate

$$
math4_i = \beta_0 + \beta_1 exppp_i + u_i
$$

using `lm()` and save the fitted model as `prob05_model`.

1. Check that the estimated intercept and slope match your calculations in Problem 4 (small differences from rounding are acceptable).
2. Interpret the slope in context. What would be the predicted change in `math4` associated with a $500 increase in expenditures per pupil?

Save the written responses to parts 2 and 3 as a string to `prob05`.

**Answer:**

```{r, problem05}
# prob05_model <-
# prob05 <-
```

---

**#6**  

Using `prob05_model`:

1. Calculate the fitted values and residuals; save them as `yhat` and `uhat`.
2. Check numerically that the sample mean of the residuals is approximately zero.
3. Identify the school with the largest absolute residual. Report its observed value for `math4`, fitted value for `math4`, residual, and value of `exppp`.
4. Explain what the sign of that residual means.

Save the requested school information as `prob06_school` and your explanation as `prob06`.

**Answer:**

```{r, problem06}
# yhat <-
# uhat <-
# prob06_school <-
# prob06 <-
```

---

**#7** 

For `prob05_model`, calculate the following manually from `math4`, `yhat`, and `uhat`:

$$
TSS=\sum_i(y_i-\bar{y})^2, \qquad
ESS=\sum_i(\hat{y}_i-\bar{y})^2, \qquad
SSR=\sum_i\hat{u}_i^2,
$$

$$
R^2=1-\frac{SSR}{TSS}, \qquad
SER=\sqrt{\frac{SSR}{n-2}}.
$$

1. Save these five values in a named vector called `prob07_fit`. Put them in the order: $TSS, ESS, SSR, R^2, SER$.
2. Interpret $R^2$ and the SER in context. Be precise about their units. Make sure your $R^2$ matches the value reported by your model `summary()`.

Save your interpretation as `prob07`.

**Answer:**

```{r, problem07}
# prob07_fit <-
# prob07 <-
```

---

**#8**  

Estimate the level-log model

$$
math4_i=\beta_0+\beta_1\ln(exppp_i)+u_i
$$

and save it as `prob08_model`.

1. Interpret the estimated slope as a semi-elasticity.
2. Compute the predicted change in `math4` for a 10% increase in spending using the exact log change $\ln(1.10)$; save it as `prob08_change`.
3. Compare the $R^2$ and residual standard error of the two models. Which functional form do you prefer for prediction, and why?

Save your written response as `prob08`.

**Answer:**

```{r, problem08}
# prob08_model <-
# prob08_change <-
# prob08 <-
```

---

**#9**  

Suppose the goal is to estimate the **causal effect** of per-pupil spending on fourth-grade math performance using the simple regression in Problem 5.

In 6–8 sentences:

1. State the conditional mean assumption needed for a causal interpretation.
2. Explain why the percentage of students eligible for free or reduced-price lunch (`lunch`) might violate this assumption.
3. Explain the likely direction of the resulting omitted-variable bias if high-poverty schools tend to have both higher spending needs and lower test scores for reasons not caused by spending.
4. Distinguish a regression that is useful for prediction from one that identifies a causal effect.

Save your response as a string named `prob09`.

**Answer:**

```{r, problem09}
# prob09 <-
```

---

**#10** . 

The OLS slope can be written as

$$
\hat\beta_1=\beta_1+
\frac{\sum_i(x_i-\bar{x})u_i}{\sum_i(x_i-\bar{x})^2}.
$$

In 6–8 sentences:

1. Use this expression to explain why $E(u_i\mid X_i)=0$ implies that $\hat\beta_1$ is unbiased.
2. Explain why $\hat\beta_1$ varies from sample to sample even when $\beta_1$ is fixed.
3. Explain how increasing the sample size generally affects the sampling distribution of $\hat\beta_1$.
4. Distinguish unbiasedness from consistency.

Save your response as a string named `prob10`.

**Answer:**

```{r, problem10}
# prob10 <-
```

